RegressionPvNP

### 2017-06-08 RegressionPvNP.Rmd
More analysis

load("Test_patients.Rdata", verbose=T)
validation.patient.data = patient.data

@@

validation.patient.info = read.patient.info(patient.file, set='Test')
sum.validation.info = summarise.patient.info(validation.patient.info)

patient.info = arrange(patient.info, Status, Patient, Endoscopy.Year, Pathology)

sum.patient.data = summarise.patient.info(patient.info)

@@

patient.data[which(sapply(patient.data, function(df) nrow(df$seg.vals)) == 0)] = NULL
sum.patient.data = as.data.frame(subset(sum.patient.data, Patient %in% names(patient.data))) ## For now

@@

pander(sum.patient.data[,c('Patient', 'Status', 'start.year', 'end.year','total.samples','total.endos','highest.path')], justify='left')

@@

```{r tile, echo=F, message=F, warning=F, include=F}
tile.w=1e07
#tile.w=5e06
tile.files = grep(as.character(tile.w), analysis.files, value=T, fixed=T)
load(tile.files[1], verbose=T)
dim(mergedDf)

@@

dysplasia.df = t(mergedDf[,names(labels)])
dim(dysplasia.df)
```

@@

crossvalidate.by.patient<-function(x,y,lambda,pts,a=1,nfolds=10, splits=5, fit=NULL, minR=0.2, select='classification') {
  if (nfolds > 5) minR = 0.1
  message(paste("Running", splits, "splits",nfolds,"times on", paste(dim(x), collapse=':'), 'alpha=',a ))
  fit.e = list()
  tpts = create.patient.sets(pts, nfolds, splits, minR)
  cv.pred = (matrix(nrow=0, ncol=length(lambda)))
  cv.binomial.deviance = (matrix(nrow=0, ncol=length(lambda)))
  for (n in 1:nfolds) {  
    message(paste(n, "fold"))
    setCol = grep(paste('^fold.',n,'$',sep=''), colnames(tpts))

    cv.class = matrix(nrow=splits, ncol=length(lambda))
    deviance = matrix(nrow=splits, ncol=length(lambda))
    for (i in 1:splits) { 
      message(paste(i, "split"))
      test.rows = which(rownames(x) %in% tpts[which(tpts[,setCol] == i), 'Samplename'])
      test = x[test.rows,]
      training = x[-test.rows,]
      # pre-spec lambda seq
      fitCV <- glmnet(training, y[-test.rows], lambda=lambda, family='binomial', alpha=a) 
      # autoplot(fit) + theme(legend.position="none")

      # Confusion matrix: quantitiative, pos results + neg results / number of test rows
      pred <- pi.hat(predict(fitCV, test, type='link')) 
      cv.class[i,] = apply(pred, 2, function(p) {
        (p%*%y[test.rows] + (1-p) %*% (1-y[test.rows]))/length(test.rows)
      })

@@

    arrange(subset(df, mean < se1 & !lambda.min & log.lambda > subset(df, lambda.min)$log.lambda ), -mean)[1:10,]
    df[arrange(subset(df, mean < se1 & !lambda.min & log.lambda > subset(df, lambda.min)$log.lambda ), -mean)[1, 'lambda.at'], 'lambda.1se'] = T
    #df[df$log.lambda < subset(df, mean == max(mean))$log.lambda-sd(df$log.lambda),][1, 'lambda-1se'] = TRUE
  } else if (select == 'deviance') {
    df$lambda.min = df$mean.b.dev == min(df$mean.b.dev)
    se1 = df[df$lambda.min == T, 'mean.b.dev'] - df[df$lambda.min == T, 'sme.b.dev']
    df[arrange(subset(df, mean.b.dev <= se1 & !lambda.min), -mean.b.dev)[1, 'lambda.at'], 'lambda.1se'] = T
    #df[df$mean.b.dev >= df[df$lambda.min == T, 'mean.b.dev'] + sd(df$mean.b.dev), ][1, 'lambda.1se'] = T
  }

@@

    l = fit0$lambda
    if (a == 0) {
      l = sort(c(fit0$lambda, seq(exp(-5), exp(-10), -1e-6),
            seq(exp(-10), exp(-15), -1e-8),
            seq(exp(-15), exp(-20), -1e-9)), decreasing=T)
    }
    cv.patient = crossvalidate.by.patient(x=dysplasia.df, y=labels, lambda=l, pts=pts, a=a, nfolds=folds, splits=splits, fit=fit0)

@@

```{r xvalpt2, message=F, warning=F, echo=T}
performance.at.1se = do.call(rbind, performance.at.1se)
performance.at.1se$alpha = rownames(performance.at.1se)

@@

cfs = as.data.frame(matrix(data=0,nrow=length(unique(names(table(unlist(lapply(coef.stable, names)))))), ncol=length(names(coef.stable)), 
                           dimnames=list( unique(names(table(unlist(lapply(coef.stable, names))))), names(coef.stable) )))

table(rowSums(cfs))

@@

stability = do.call(rbind, lapply(coef.stable, function(cf) {
  cbind('75%'=length(which(cf >= 0.75)),  '50%'=length(which(cf >= 0.5)),  '<50%'=length(which(cf < 0.5)))
}))
rownames(stability) = names(coef.stable)

pander(as.data.frame(sort(names(which((coef.stable[['1']]) >= 0.75)))), justify='left')
```

@@

@@

`r pander(table(rowSums(cfs)), caption="Number of features found in 1, 2, or all 3 models by alpha 0.8, 0.9 and 1")`

@@

### Features in samples

```{r echo=F, warning=F, message=F}

length(which( coef.stable[['1']] >= 0.5) )

regions = rownames(coefs[['1']])

@@

#which(meantests < 0.05)

summary(pg[,'1:119640299-129610322'])
summary(nonpg[,'1:119640299-129610322'])

@@

```{r, warning=F, message=F}
#topF = names(which((coef.stable[['1']]) >= 0.75))
topF = names(coef.stable[['1']])

@@

# Gains/losses enriched in the top features?
fisher.test(rbind(colSums(gl[,c('P.gain','P.loss')]),colSums(allgl[-feat,c('ALL.P.gain','ALL.P.loss')])  ))

fisher.test(rbind(colSums(gl[,c('NP.gain','NP.loss')]),colSums(allgl[-feat,c('ALL.NP.gain','ALL.NP.loss')])  ))

@@

# Number of segments is strongly correlated with the length of the chromosome
with(topFRegions, cor.test(length, n.segments))

topFRegions$adj.info.content = topFRegions$protein_coding/(topFRegions$length^0.6)
with(topFRegions, cor.test(length, adj.info.content))

topFRegions = transform(topFRegions, ai.cut=cut(adj.info.content, 5, labels=c(5:1), include.lowest=T))           # bin data
ggplot(topFRegions, aes(reorder(chr, -adj.info.content), totalbp/length, fill=ai.cut)) + ylim(0, 0.5) +
  labs(y='Feature bp/total chr length', title='Ratio of covered base pairs', x='chr') +
  geom_bar(stat='identity') +  geom_text(aes(label=round(totalbp/length, 2)), vjust=-0.5) +
  scale_fill_discrete(name='Length adjusted information content') + theme(legend.position = 'bottom')

@@

@@

Strong correlation between the length adjusted information content (protein coding genes) and the covered bases in a chromosome r=`r round(ct$estimate, 2)` (p-value `r round(ct$p.value, 4)`). No features from chromosome 19 (the most gene dense chromosome) were selected.

`r round(bpCovered, 2)*100`% of the total bases are within the top selected features.

@@

bm = getBM(mart=biomart, attributes=attr, 
           filters=list('chromosomal_region'=sub('-', ':',topF), "biotype"=c("IG_C_gene","IG_D_gene","IG_J_gene","IG_V_gene","protein_coding","TR_C_gene","TR_D_gene","TR_J_gene","TR_V_gene")))
#head(bm)
bm = bm[-which(bm$hgnc_symbol == ""),]
bm = bm %>% group_by(hgnc_symbol, chromosome_name, start_position, end_position) %>% summarise('exon_count' = length(ensembl_exon_id)) 
exons = bm %>% group_by(chromosome_name) %>% summarise('exons'=sum(exon_count))

@@

# Not related to the length of the segment
with(topFSegments, cor.test(genes, (end-start)))

head(bm)

@@

with(topFRegions, cor.test(protein_coding, exon.cnt))
with(topFRegions, cor.test(gene.cnt, exon.cnt)) # expected

# Exons correlate with the adjusted information content measure
with(topFRegions, cor.test(exon.cnt, adj.info.content))

ggplot(topFRegions, aes(reorder(chr, -adj.info.content), gene.cnt/protein_coding, fill=ai.cut)) + geom_bar(stat='identity') +
  labs(y='Gene Ratio', x='chr', title='Feature genes vs all protein coding genes') +
  geom_text(aes(label=round(gene.cnt/protein_coding, 2)), vjust=-0.5) +
  scale_fill_discrete(name='Length adjusted information content') + theme(legend.position = 'bottom')

@@

@@

ggplot(topFRegions, aes(reorder(chr, -adj.info.content), COSMIC.cont/COSMIC.all, fill=ai.cut)) + geom_bar(stat='identity') +
  labs(y='COSMIC Gene Ratio', x='chr', title='COSMIC genes: Feature vs all')+
  geom_text(aes(label=round(COSMIC.cont/COSMIC.all, 2)), vjust=-0.5) +
  scale_fill_discrete(name='Length adjusted information content') + theme(legend.position = 'bottom')
```

@@

@@

stability = do.call(rbind, lapply(coef.stable, function(cf) {
  cbind('75%'=length(which(cf >= 0.75)),  '50%'=length(which(cf >= 0.5)),  '<50%'=length(which(cf < 0.5)))
}))
rownames(stability) = names(coef.stable)
```

@@

`r pander(table(rowSums(cfs)), justify='left')`

@@

stability = do.call(rbind, lapply(coef.stable, function(cf) {
  cbind('75%'=length(which(cf >= 0.75)),  '50%'=length(which(cf >= 0.5)),  '<50%'=length(which(cf < 0.5)))
}))
rownames(stability) = names(coef.stable)

@@

```{r, echo=T, message=F, warning=F, fig.height=16, fig.width=16}
m = median(subset(sum.patient.data, Status == 'P')$total.samples)
nonps = as.vector(unlist(lapply(patient.data[subset(sum.patient.data, Status == 'NP')$Patient], function(df) df$info$Samplename)))

file = paste(cache.dir, 'few.Rdata', sep='/')

if (file.exists(file)) {
  load(file, verbose=T)
} else {
  # Few
  samples = c(nonps, as.vector(unlist(lapply(patient.data[subset(sum.patient.data, Status == 'P' & total.samples < m)$Patient], function(pt) pt$info$Samplename))))

@@

```{r, echo=F, warning=F, message=F, fig.height=8, fig.width=8, eval=F}
plot.endo( subset(endos, Patient %in% subset(me, max.endo <= 2 & max.endo > 0)$Patient), min(endos$months.before.final), max(endos$months.before.final)  )
```

```{r, echo=F, warning=F, message=F, fig.height=4, fig.width=4, eval=F}
plot.endo( subset(endos, Patient %in% subset(me, max.endo == 0)$Patient),min(endos$months.before.final), max(endos$months.before.final) )
```

@@

```{r, echo=T, warning=F, message=F, eval=F}
all = do.call(rbind, pg.samp)

@@

## Similar numbers of each in these batches so how did we do?
tfPR = sapply( c('Exome_subcohort','Batch_1','Batch_2','Batch_3','Batch_4'), function(b) {
  pred = subset(prog, Batch.Name == b)$Prediction
  predB = subset(nonp, Batch.Name == b)$Prediction

  tpr = length(which(pred > 0.5))/length(pred)
  fpr = length(which(predB > 0.5))/length(predB)
  cbind.data.frame('TPR'=tpr, 'FPR'=fpr)
})

tfPR

## Exome_cohort fpr driven by the two patients that we predicted to be progressors in nearly every sample
length(which(pg.samp$AD0591$Prediction > 0.5))
length(which(pg.samp$AH0329$Prediction > 0.5))

## Except the exome cohort, the TPR rate appears to be right around 48-68%
## FPR is over 20% in Batch_4 though

unique(subset(nonp, Batch.Name == 'Batch_4')$Patient)

pg.samp$AHM0100$Prediction > 0.5
pg.samp$AHM0121$Prediction > 0.5
pg.samp$AD0644$Prediction > 0.5

cb = combn( as.character(unique(all$Batch.Name)), 2)

batch =  sapply(1:ncol(cb), function(i) {
  sapply(c('SampleSD','SampleMEAN', 'Total.Reads'), function(x) {
    tt = t.test(subset(all, Batch.Name == cb[1,i])[[x]], subset(all, Batch.Name == cb[2,i])[[x]])  
    tt$p.value
  })
}) 
```

```{r, echo=T, warning=F, message=F, eval=F}
smp<-function(df) {
  df[,c('Prediction', 'Prediction.Dev.Resid', 'SampleMEAN', 'SampleSD', 'Barretts.Cellularity', 'Total.Reads', 'p53.Status', 'PID')]
}

prog = do.call(rbind, lapply(pg.samp[subset(sum.patient.data, Status == 'P')$Patient], smp) )
nonp = do.call(rbind, lapply(pg.samp[subset(sum.patient.data, Status == 'NP')$Patient], smp) )

GGally::ggcorr(data=prog, method=c('pairwise','pearson'), label=T) + labs(title='Pearson correlation for several QC measures')

# Compare poorly predicted to well predicted based on the sample SD measures
ttfunc<-function(col, df=NULL) {
  tt = t.test(subset(df, Prediction < 0.5)[[col]], subset(df, Prediction > 0.5)[[col]])
  cbind.data.frame('p.val'=tt$p.value, 'statistic'=tt$statistic)
}

@@

```{r, echo=T, warning=F, message=F, eval=F}
 pander(sapply(colnames(prog)[-(1:2)], ttfunc, df=prog), caption='Progressor t-test stats for QC measures between predictions probabilies  >&< 0.5 ')

 pander(sapply(colnames(nonp)[-(1:2)], ttfunc, df=nonp), caption='Non-progressor t-test stats for QC measures between predictions probabilies  >&< 0.5 ')

hist(prog$SampleSD)

subset(prog, SampleSD <= 0.025)

@@

>>>>

load("Test_patients.Rdata", verbose=T)
validation.patient.data = patient.data
length(validation.patient.data)

@@

validation.patient.info = read.patient.info(patient.file, set='Test')
validation.patient.info = arrange(validation.patient.info, Status, Patient, Endoscopy.Year, Pathology)
sum.validation.info = summarise.patient.info(validation.patient.info)

patient.info = arrange(patient.info, Status, Patient, Endoscopy.Year, Pathology)
sum.patient.data = summarise.patient.info(patient.info)

@@

if (length(which(sapply(validation.patient.data, function(df) nrow(df$seg.vals)) == 0)) > 0)
  warning(paste("Patients missing data:",
                paste(names(which(sapply(patient.data, function(df) nrow(df$seg.vals)) == 0)), collapse=', ')))

nor = which(sapply(patient.data, function(df) nrow(df$seg.vals)) == 0)
if (length(nor) > 0) patient.data[nor] = NULL

nor = which(sapply(validation.patient.data, function(df) nrow(df$seg.vals)) == 0)
if (length(nor) > 0) validation.patient.data[nor] = NULL

sum.patient.data = as.data.frame(subset(sum.patient.data, Patient %in% names(patient.data))) ## For now

@@

for (pt in names(validation.patient.data)) {
  #print(pt)
  #print(patient.data[[pt]]$info$Samplename %in% colnames(patient.data[[pt]]$seg.vals)[-(1:5)])
  validation.patient.data[[pt]]$info = validation.patient.data[[pt]]$info[validation.patient.data[[pt]]$info$Samplename %in% colnames(validation.patient.data[[pt]]$seg.vals)[-(1:5)],]
}

pander(sum.patient.data[,c('Patient', 'Status', 'start.year', 'end.year','total.samples','total.endos','highest.path')], justify='left', caption='Training patient set')

pander(sum.validation.info[,c('Patient', 'Status', 'start.year', 'end.year','total.samples','total.endos','highest.path')], justify='left', caption='Validation patient set')

@@

```{r tile, echo=F, message=F, warning=F, include=F}
#tile.w=1e07
tile.w=5e06
tile.files = grep(as.character(tile.w), analysis.files, value=T, fixed=T)
load(tile.files[grep('Test|validation', tile.files)], verbose=T)
mergedDfValidation = mergedDf
dim(mergedDfValidation)

load(tile.files[grep('Test|validation', tile.files, invert=T)], verbose=T)
dim(mergedDf)

@@

dysplasia.df = t(mergedDf[,names(labels)])
dim(dysplasia.df)

validation.df = t(mergedDfValidation[, grep('^D', colnames(mergedDfValidation))])
dim(validation.df)
```

@@

crossvalidate.by.patient<-function(x,y,lambda,pts,a=1,nfolds=10, splits=5, fit=NULL, minR=0.2, select='deviance') {
  if (nfolds > 5) minR = 0.1
  x = as.matrix(x)
  message(paste("Running", splits, "splits",nfolds,"times on", paste(dim(x), collapse=':'), 'alpha=',a ))
  fit.e = list()
  tpts = create.patient.sets(pts, nfolds, splits, minR)
  cv.pred = (matrix(nrow=0, ncol=length(lambda)))
  cv.binomial.deviance = (matrix(nrow=0, ncol=length(lambda)))
  for (n in 1:nfolds) {  
    message(paste(n, "fold"))
    setCol = grep(paste('^fold.',n,'$',sep=''), colnames(tpts))

    cv.class = matrix(nrow=splits, ncol=length(lambda))
    deviance = matrix(nrow=splits, ncol=length(lambda))
    for (i in 1:splits) { 
      message(paste(i, "split"))
      test.rows = which(rownames(x) %in% tpts[which(tpts[,setCol] == i), 'Samplename'])
      test = x[test.rows,]
      training = x[-test.rows,]
      # pre-spec lambda seq
      fitCV <- glmnet(training, y[-test.rows], lambda=lambda, alpha=a, family='binomial') 
      # autoplot(fit) + theme(legend.position="none")

      # Confusion matrix: quantitiative, pos results + neg results / number of test rows
      pred <- pi.hat(predict(fitCV, test, type='link'))  # pi.hat turns these into probabilities from logit
      cv.class[i,] = apply(pred, 2, function(p) {
        (p%*%y[test.rows] + (1-p) %*% (1-y[test.rows]))/length(test.rows)
      })

@@

    lambda.1se.search = subset(df, mean < se1 & !lambda.min & log.lambda > subset(df, lambda.min)$log.lambda) 
    if (nrow(lambda.1se.search) > 0) {
      arrange(lambda.1se.search, -mean)[1:10,]
      df[arrange(subset(df, mean < se1 & !lambda.min & log.lambda > subset(df, lambda.min)$log.lambda ), -mean)[1, 'lambda.at'], 'lambda.1se'] = T
    }
    #df[df$log.lambda < subset(df, mean == max(mean))$log.lambda-sd(df$log.lambda),][1, 'lambda-1se'] = TRUE
  } else if (select == 'deviance') {
    df[which.min(df$mean.b.dev), 'lambda.min'] = T

    lmin = subset(df, lambda.min)$lambda
    df[subset(df, log.lambda <= log(lmin) + log(sd(lambda))/2 )[1, 'lambda.at'], 'lambda.1se'] = T
  }

@@

    l = fit0$lambda
    if (a == 0) {
      l = sort(c(fit0$lambda, seq(exp(-5), exp(-10), -1e-6),
            seq(exp(-10), exp(-15), -1e-8),
            seq(exp(-15), exp(-20), -1e-9)), decreasing=T)
    }
    cv.patient = crossvalidate.by.patient(x=dysplasia.df, y=labels, lambda=l, pts=pts, a=a, nfolds=folds, splits=splits, fit=fit0, select='deviance')

@@

```{r xvalpt2, message=F, warning=F, echo=T}
select.alpha = as.character(0.9)

performance.at.1se = do.call(rbind, performance.at.1se)
performance.at.1se$alpha = rownames(performance.at.1se)

@@

cfs = as.data.frame(matrix(data=0,nrow=length(unique(names(table(unlist(lapply(coef.stable, names)))))), ncol=length(names(coef.stable)), 
                           dimnames=list( unique(names(table(unlist(lapply(coef.stable, names))))), names(coef.stable) )))
for (i in names(coef.stable)) 
  cfs[intersect(rownames(cfs), names(coef.stable[[i]])), i] = 1

@@

stability = do.call(rbind, lapply(coef.stable, function(cf) {
  cbind('>=75%'=length(which(cf >= 0.75)),  '>=50%'=length(which(cf >= 0.5)),  '>=25%'=length(which(cf >= 0.25)), 'Total'=length(cf))
}))
rownames(stability) = names(coef.stable)

pander(stability, justify='left', caption='Number of features stable in n% or more folds')
```

@@

### Validation Data

TBD

```{r, message=F, warning=F, echo=F, eval=F}

fitV <- glmnet(dysplasia.df, labels, alpha=select.alpha, nlambda=nl, family='binomial') # all patients
#autoplot(fitV, xvar='lambda', main=paste('fit0, all samples, alpha=',a,sep='')) + theme(legend.position='none') 
cv.patient = crossvalidate.by.patient(x=dysplasia.df, y=labels, lambda=fitV$lambda, pts=pts, a=select.alpha, nfolds=folds, splits=splits, fit=fitV)
lambda.opt = ifelse( length(cv.patient$lambda.1se) > 0, cv.patient$lambda.1se, cv.patient$lambda.min )

pred = predict(fitV, newx=validation.df, s=lambda.opt, type='response')

statusV = unlist(sapply(validation.patient.data, function(df) {
  df$info = arrange(df$info, Endoscopy.Year, Pathology)
  label = as.integer(df$info$Status == 'P') #as.integer(df$info$Pathology %in% c('HGD', 'IMC'))
  names(label) = df$info$Samplename
  return(label)
}))
names(statusV) = sub('.*\\.', '',  names(statusV))

# Progressors - by sample
length( which(pred[ names(which(statusV == 1)), ] > 0.5) )/length(which(statusV == 1))

# Non Progressors - by sample
length( which(pred[ names(which(statusV == 0)), ] < 0.5) )/length(which(statusV == 0))

@@

`r pander(colSums(cfs), caption="Number of features found in 1, 2, or all 3 models by alpha 0.8, 0.9 and 1. While there are fewer features for higher values of alpha, the features are the same across models.")`

@@

```{r echo=F, warning=F, message=F, eval=F}
### Features in samples
## TODO Look at the coefficients, are the high pos ones "gains" and the low neg ones "losses" etc
plot(coefs[[select.alpha]][,1])

length(which( coef.stable[[select.alpha]] >= 0.5) )

regions = rownames(coefs[[select.alpha]])

@@

which(meantests < 0.05)

#summary(pg[,'1:119640299-129610322'])
#summary(nonpg[,'1:119640299-129610322'])

@@

Using the features from the alpha=0.9 models

```{r, warning=F, message=F}
#topF = names(which((coef.stable[['1']]) >= 0.75))
topF = names(coef.stable[[select.alpha]])

@@

# Gains/losses enriched in the top features?  Appears so
(fisher.test(rbind(colSums(gl[,c('P.gain','P.loss')]),colSums(allgl[-feat,c('ALL.P.gain','ALL.P.loss')])  )) )

(fisher.test(rbind(colSums(gl[,c('NP.gain','NP.loss')]),colSums(allgl[-feat,c('ALL.NP.gain','ALL.NP.loss')])  )) )

@@

# Number of segments is strongly correlated with the length of the chromosome
#with(topFRegions, cor.test(length, n.segments))

topFRegions$adj.info.content = topFRegions$protein_coding/(topFRegions$length^0.6)
#with(topFRegions, cor.test(protein_coding, length)) # Adjust the length so that the correlation between length and

topFRegions = transform(topFRegions, ai.cut=cut(adj.info.content, 5, labels=c(5:1), include.lowest=T))           # bin data

grid.arrange(ggplot(topFRegions, aes(reorder(chr, -adj.info.content), totalbp/length, fill=ai.cut)) + ylim(0, 0.5) +
  labs(y='Feature bp/total chr length', title='Ratio of covered base pairs', x='chr') +
  geom_bar(stat='identity') +  geom_text(aes(label=round(totalbp/length, 2)), vjust=-0.5) +
  scale_fill_discrete(name='Length adjusted information content') + theme(legend.position = 'bottom'), 
  bottom='Ratio of base pair coverage per chromosome in the top non-zero features selected from the model.\nChromosomes are sorted and binned by their gene (e.g. information) content rather than by overall length.')

@@

with(topFRegions, cor.test(adj.info.content, totalbp/length))

@@

`r pander(ct, caption="Pearson correlation between the adjusted information content (genes) per chromosome, and the ratio of length of the features")`

`r round(bpCovered, 2)*100`% of the all bps in the genome are within the top selected features.

@@

bm = getBM(mart=biomart, attributes=attr, 
           filters=list('chromosomal_region'=sub('-', ':',topF), "biotype"=c("IG_C_gene","IG_D_gene","IG_J_gene","IG_V_gene","protein_coding","TR_C_gene","TR_D_gene","TR_J_gene","TR_V_gene")))
#head(bm)

bm = bm[-which(bm$hgnc_symbol == ""),]
bm = bm %>% group_by(ensembl_gene_id, hgnc_symbol, chromosome_name, start_position, end_position) %>% summarise('exon_count' = length(ensembl_exon_id)) 
exons = bm %>% group_by(chromosome_name) %>% summarise('exons'=sum(exon_count))

write.table(bm$ensembl_gene_id, sep='\t', col.names = F, row.names = F, quote = F, file='feature_genes.txt')

@@

# Not related to the length of the segment
pander(with(topFSegments, cor.test(genes, (end-start))), caption="Number of genes per segment is not related to the segment length.")

@@

#with(topFRegions, cor.test(protein_coding, exon.cnt))
#with(topFRegions, cor.test(gene.cnt, exon.cnt)) # expected

# Exons correlate with the adjusted information content measure, expected
#with(topFRegions, cor.test(exon.cnt, adj.info.content))

grid.arrange(ggplot(topFRegions, aes(reorder(chr, -adj.info.content), gene.cnt/protein_coding, fill=ai.cut)) + geom_bar(stat='identity') +
  labs(y='Gene Ratio', x='chr', title='Feature genes vs all protein coding genes') +
  geom_text(aes(label=round(gene.cnt/protein_coding, 2)), vjust=-0.5) +
  scale_fill_discrete(name='Length adjusted information content') + theme(legend.position = 'bottom'), 
  bottom='Ratio of genes found within the boundaries of the top features vs all genes on the chromosome.\nChromosomes are ordered and binned by adjusted gene (e.g. information) content rather than length.')

@@

write.table(containedCC$ensembl_gene_id, quote=F, row.names = F, col.names = F, file='feature_cosmic.txt')

@@

grid.arrange(ggplot(topFRegions, aes(reorder(chr, -adj.info.content), COSMIC.cont/COSMIC.all, fill=ai.cut)) + geom_bar(stat='identity') +
  labs(y='COSMIC Gene Ratio', x='chr', title='COSMIC genes: Feature vs all')+
  geom_text(aes(label=round(COSMIC.cont/COSMIC.all, 2)), vjust=-0.5) +
  scale_fill_discrete(name='Length adjusted information content') + theme(legend.position = 'bottom'), 
  bottom='Again, but this time looking only at COSMIC genes, the ratio of COSMIC genes\nwithin the selected feature vs all COSMIC genes on that chromosome.\nChromosomes are ordered and colored by information content.')

colSums(topFRegions[,c('COSMIC.all','COSMIC.cont')],na.rm=T)
colSums(topFRegions[,c('protein_coding','gene.cnt')],na.rm=T)

#table(topFRegions$ai.cut)

@@

### Gains vs Losses

try

gains = pmax(0, x-median(x))
losses = pmin(0, x-median(x))

```{r warning=F, message=F, fig.height=8, fig.width=8}
run.cv<-function(x, y, alphas=c(0,0.2,0.5,1), ...) {
  plots = list(); coefs = list(); performance.at.1se = list()
  for (a in alphas) {
    fitg <- glmnet(x, y, alpha=a, nlambda=nl, family='binomial') # all patients

    l = fitg$lambda
    if (a == 0) {
      l = sort(c(fitg$lambda, seq(exp(-5), exp(-10), -1e-6),
            seq(exp(-10), exp(-15), -1e-8),
            seq(exp(-15), exp(-20), -1e-9)), decreasing=T)
    }

    cvp = crossvalidate.by.patient(x, y, fit=fitg, lambda=l, a=a, ...)
    plots[[as.character(a)]] = arrangeGrob(cvp$plot+ggtitle('Classification'), cvp$deviance.plot+ggtitle('Binomial Deviance'), top=paste('alpha=',a,sep=''), ncol=2)

    coef.1se = as.data.frame(non.zero.coef(fitg, cvp$lambda.1se))
    coefs[[as.character(a)]] = coef.stability(coef.1se, cvp$non.zero.cf)

    performance.at.1se[[as.character(a)]] = subset(cvp$lambdas, lambda.1se == T)

  }
  return(list('plots'=plots, 'coefs'=coefs, 'performance'=performance.at.1se))
}

file = paste(cache.dir, 'gain.loss.Rdata', sep='/')
if (file.exists(file)) {
  message(paste("loading", file))
  load(file, verbose=T)
} else {
  gain.df = as.data.frame(matrix(nrow=nrow(dysplasia.df), ncol=ncol(dysplasia.df),data=0, dimnames=list(rownames(dysplasia.df), colnames(dysplasia.df))))
  loss.df = gain.df

  for (i in 1:nrow(dysplasia.df)) {
     #gain.df[i,] = as.integer(dysplasia.df[i,] >= 1.1)
     #loss.df[i,] = as.integer(dysplasia.df[i,] <= 0.9)
     #all.binary[i,] = as.integer(dysplasia.df[i,] >= 1.1 | dysplasia.df[i,] <= 0.9)
    gain.df[i,] = pmax(0, dysplasia.df[i,] - median(dysplasia.df[i,]))
    loss.df[i,] = pmin(0, dysplasia.df[i,] - median(dysplasia.df[i,]))
  } 
  #hist(colSums(gain.df))
  #hist(colSums(loss.df))
  #hist(colSums(all.binary))

  gp = run.cv(x=as.matrix(gain.df), y=labels, pts=pts, nfolds=folds, splits=splits, alphas=c(0,0.5,0.8,1))
  lp = run.cv(x=as.matrix(loss.df), y=labels, pts=pts, nfolds=folds, splits=splits, alphas=c(0,0.5,0.8,1))

  colnames(gain.df) = paste(colnames(gain.df), '_GAIN', sep='')
  colnames(loss.df) = paste(colnames(loss.df), '_LOSS', sep='')

  gl.df = merge(gain.df, loss.df, by='row.names')
  rownames(gl.df) = gl.df$Row.names
  gl.df$Row.names = NULL

  both = run.cv(x=as.matrix(gl.df), y=labels, pts=pts, nfolds=folds, splits=splits, alphas=c(0,0.5,0.8,1))

  save(gp, lp, all, file=file)
}

do.call(grid.arrange, c(gp$plots, top='All samples, GAINS 10fold, 5 splits'))
do.call(grid.arrange, c(lp$plots, top='All samples, LOSSES 10fold, 5 splits'))
do.call(grid.arrange, c(both$plots, top='All samples, gains & losses 10fold, 5 splits'))

@@

stability = do.call(rbind, lapply(coef.stable, function(cf) {
  cbind('>=75%'=length(which(cf >= 0.75)),  '>=50%'=length(which(cf >= 0.5)),  '>=50%'=length(which(cf >= 0.25)), 'Total'=length(cf))
}))
rownames(stability) = names(coef.stable)
```

@@

@@

stability = do.call(rbind, lapply(coef.stable, function(cf) {
  cbind('>=75%'=length(which(cf >= 0.75)),  '>=50%'=length(which(cf >= 0.5)),  '>=50%'=length(which(cf >= 0.25)), 'Total'=length(cf))
}))
rownames(stability) = names(coef.stable)

@@

```{r manyfew, echo=T, message=F, warning=F, fig.height=16, fig.width=16}
m = median(subset(sum.patient.data, Status == 'P')$total.samples)
nonps = as.vector(unlist(lapply(patient.data[subset(sum.patient.data, Status == 'NP')$Patient], function(df) df$info$Samplename)))

file = paste(cache.dir, 'few.Rdata', sep='/')
if (file.exists(file)) {
  load(file, verbose=T)
} else {
  # Few
  samples = c(nonps, as.vector(unlist(lapply(patient.data[subset(sum.patient.data, Status == 'P' & total.samples < m)$Patient], function(pt) pt$info$Samplename))))

@@

```{r, echo=F, warning=F, message=F, fig.height=6, fig.width=4}
plot.endo( subset(endos, Patient %in% subset(me, max.endo <= 2 & max.endo > 0)$Patient), min(endos$months.before.final), max(endos$months.before.final)  )
```

```{r, echo=F, warning=F, message=F, fig.height=4, fig.width=4}
plot.endo( subset(endos, Patient %in% subset(me, max.endo == 0)$Patient),min(endos$months.before.final), max(endos$months.before.final) )
```

@@

```{r dq_batch, echo=T, warning=F, message=F}
all = do.call(rbind, pg.samp)

@@

## Similar numbers of each in these batches so how did we do?
tfPR = sapply(as.character(unique(patient.info$Batch.Name)), function(b) {
  pred = subset(prog, Batch.Name == b)$Prediction
  predB = subset(nonp, Batch.Name == b)$Prediction

  # Only for progressors because we know for certain they progress, the non-ps may just not have progressed *yet*
  tpr = length(which(pred > 0.5))/length(pred)
  fpr = length(which(predB > 0.5))/length(predB)
  cbind.data.frame('TPR'=tpr, 'FPR'=fpr)
})
```

`r pander(t(tfPR), justify='left', caption=paste('True positive rate is between', paste(round(range(tfPR['TPR',], na.rm=T), 2), collapse='-')))`

```{r dq_batch2, echo=F, warning=F, message=F}

pilot = lapply(pg.samp, function(df) subset(df, Batch.Name == 'NP_Pilot_Study'))
pilot = do.call(rbind, pilot[ which(sapply(pilot, nrow) > 0) ])
pilot = subset(pilot, Status == 'NP')
length(which(pilot$Prediction > 0.5))/nrow(pilot)

p2 = subset(pilot, Patient %nin% c('AD0591', 'AH0329'))
length(which(p2$Prediction > 0.5))/nrow(p2)

exome = lapply(pg.samp, function(df) subset(df, Batch.Name == 'Exome_subcohort'))
exome = do.call(rbind, exome[ which(sapply(exome, nrow) > 0) ])
exome = subset(exome, Status == 'NP')
length(which(exome$Prediction > 0.5))/nrow(exome)

e2 = subset(exome, Patient %nin% c('AD0591', 'AH0329'))
length(which(e2$Prediction > 0.5))/nrow(e2)

@@

The false positive rate in both the exome and NP_pilot batches is driven by two patients that we predict to be progressors in nearly every sample: AD0591 and AH0329.  Both patients had samples in the exome batch, while AH0329 also had samples in the NP_Pilot batch.  Without those patients the FPR in the exome batch was `r round(length(which(e2$Prediction > 0.5))/nrow(e2), 2)*100`% for `r length(unique(e2$Patient))` patient(s), and in the NP_Pilot batch the FPR was `r round(length(which(p2$Prediction > 0.5))/nrow(p2), 2)*100`% for `r length(unique(p2$Patient))` patient(s).

The somewhat lower Batch_1 prediction rate is driven by the samples from AHM0277 as only `r length(which(pg.samp$AHM0277$Prediction > 0.5))/nrow(pg.samp$AHM0277)*100`% of this patient's samples were predicted to progress.  Batch_3 is driven by three other patients that have low prediction rates: PR1_ADH_076, PR1_HIN_046, and PR1_WSH_026.

### Look at these specific patients

```{r, echo=F, eval=F}

pg.samp$AD0591
pg.samp$AH0329
pg.samp$PR1_ADH_076
pg.samp$PR1_HIN_046
pg.samp$PR1_WSH_026

@@

## Cellularity or general data quality measures?

```{r, echo=F, warning=F, message=F}
all$SampleCoV = with(all, SampleSD/SampleMEAN)
